---
title: 
  !expr eval(10*5)
output: 
  html_document
---

```{r}
mtcars
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval=T}
here::here()
```
Here we develop an ecological model to assess whether strains which resist beneficial FGEs can increase in frequency in populations. 

We consider two types of foreign genetic elements (FGEs): the first FGE is denoted as $G_{1}$ is modestly beneficial as it provides protection against a single antibiotic. The second FGE is denoted as $G_{2}$ and is highly beneficial and provides protection against two antibiotics. 

We then assess the ecological dynamics of a 'choosy' strain (denoted by subscript $c$) that resists infection by the good FGE $G_{1}$ with probability $\pi_{G_{1}}$, whereas it has no resistance against infection by the better FGE $\pi_{G_{2}} = 0$. Here, resistance reflects the probability that an FGE of type $G_{i}$ is detected and degraded by CRISPR (see also @Gandon2013).

The choosy strain is in competition with a promiscuous strain (denoted by subscript $p$) that has no resistance against either type of FGE. Let $S_{p}$ and $S_{c}$ then be the number of susceptible individuals of the promiscuous and choosy strain respectively, whereas $I_{pG_{i}}$ and $I_{cG_{i}}$ are the number of individuals of the promiscuous and choosy strain infected with an FGE of type $i$ where $i \in (1,2)$. The change in numbers of individuals is then described by the following system of differential equations:
\begin{align}
\frac{\mathrm{d}S_{c}}{\mathrm{d}t} &= b_{S} \left ( \pi_{G_{1}} \right ) \left(1 - \kappa N \right ) S_{c} + \gamma_{G_{1}} I_{G_{1}} + \gamma_{G_{2}} I_{G_{2}} - \left [ d + 
  \left (1-\pi_{G_{1}} \right ) \psi_{G_{1}} + \psi_{G_{2}} \right ] S_{c} \\
  %
\frac{\mathrm{d}S_{p}}{\mathrm{d}t} &= b_{S} \left (0 \right ) \left(1 - \kappa N \right ) S_{p} + \gamma_{G_{1}} I_{G_{1}} + \gamma_{G_{2}} I_{G_{2}} - \left [ d + \psi_{G_{1}} + \psi_{G_{2}} \right ] S_{p} \\
  %
\frac{\mathrm{d}I_{p G_{i}}}{\mathrm{d}t} &= b_{G_{i}} \left ( 1 - \kappa N \right ) I_{p G_{i}} + \psi_{G_{i}} S_{p} - \left [ d_{G_{i}} + \gamma_{G_{i}} \right ] I_{G_{i}} \\
%
\frac{\mathrm{d}I_{c G_{1}}}{\mathrm{d}t} &= b_{G_{1}} \left ( 1 - \kappa N \right ) I_{c G_{1}} + \left ( 1 - \pi_{G_{1}}\right ) \psi_{G_{1}} S_{c} - \left [ d_{G_{1}} + \gamma_{G_{1}} \right ] I_{c G_{1}} \\
%
\frac{\mathrm{d}I_{c G_{2}}}{\mathrm{d}t} &= b_{G_{2}} \left ( 1 - \kappa N \right ) I_{c G_{2}} +  \psi_{G_{2}} S_{c} - \left [ d_{G_{1}} + \gamma_{G_{1}} \right ] I_{c G_{1}} 
\end{align}
where $N$ is the total population size, $N = \sum_{k \in (c,p)} S_{k} +\sum_{k \in (c,p)} \sum_{i \in (1,2)} I_{k G_{i}}$. Susceptible individuals reproduce at rate $b \left(x \right )$, where
\begin{align}
b_{y} \left ( x \right ) = 
\end{align}
and when infected  allowing for lysogenic transmission
 
## References
